Grove: A Convergent Collaborative Structure-Editor Calculus

### 2020-07-13 grove-paper.tex
Updates from Overleaf

@@

\subsection{Scenario X: Holes in the Tree}
\label{sub:Holes in the Tree}

\begin{figure}
\centering 
\includegraphics[scale=0.75]{graphviz/simple-tree-with-hole.dot}
\caption{Scenario X: Holes in the Tree}
\label{fig:Scenario X: Holes in the Tree}
\end{figure}

DIAGRAM: 2 + HOLE

NEEDS UI EXAMPLE

\subsection{Scenario A: Editing Different Parts of the Code}

@@

\begin{figure}
\centering
\includegraphics[scale=0.75]{graphviz/editing-different-parts-of-the-code-1.dot}
\includegraphics[scale=0.75]{graphviz/editing-different-parts-of-the-code-2.dot}
\includegraphics[scale=0.75]{graphviz/editing-different-parts-of-the-code-3.dot}
\includegraphics[scale=0.75]{graphviz/editing-different-parts-of-the-code-4.dot}
\caption{Scenario A: Editing Different Parts of the Code}
\label{fig:Scenario A: Editing Different Parts of the Code}
\end{figure}

\subsection{Scenario B: Edits to Nested Parts of the Code}

\begin{figure}
\centering
\begin{tikzpicture}
% TODO: texttt
% TODO: position the different graphs
\path (0cm,0cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    * -> {
      2,
      x
    }
  }
};
\path (-3cm,-2cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    + -> {
      1,
      * -> {
        2,
        x
      }
    }
  }
};
\path (3cm,-2cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    * -> {
      3,
      x
    }
  }
};
\path (0cm,-4cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    + -> {
      1,
      * -> {
        3,
        x
      }
    }
  }
};
\end{tikzpicture}
\caption{Scenario B: Edits to Nested Parts of the Code}
\label{fig:Schenario-B}
\end{figure}

TODO: REDO to MOVE a SUBTREE

A edits part of the code

B moves the code surrounding the edits of A to another part of the code

@@

\subsection{Scenario C: Direct Conflicts}

\begin{figure}
\centering
\begin{tikzpicture}
% TODO: texttt
% TODO: position the different graphs
\path (0cm,0cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    x
  }
};
\path (-3cm,-2cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    + -> {
      1,
      x
    }
  }
};
\path (3cm,-2cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    * -> {
      x,
      2
    }
  }
};
\path (0cm,-4cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    + -> {
      1,
      x
    },
    * -> {
      x,
      2
    }
  }
};
\path (2cm,-8cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    + -> {
      1,
      * -> {
        x,
        2
      }
    }
  }
};
\path (-2cm,-8cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    * -> {
      + -> {
        1,
        x
      },
      2
    }
  }
};
\end{tikzpicture}
\caption{Scenario C: Direct Conflicts}
\label{fig:Scenario-C}
\end{figure}

DIAGRAM: lam x. x

DIAGRAM: lam x. 1 + x

DIAGRAM: lam x. 2 * x

DIAGRAM: Merge: diamond with "1 +" and "2 *"

@@

UI Level Move = Delete and Add

DIAGRAM: Resolution \#1: lam x. 1 + 2 * x

DIAGRAM: Resolution \#2: lam x. 2 * (1 + x)

@@

\subsection{Scenario D: Deletion of Code Being Edited}

\begin{figure}
\centering
\begin{tikzpicture}
% TODO: texttt
% TODO: position the different graphs
\path (0cm,0cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    + -> {
      2,
      x
    }
  }
};
\path (-3cm,-2cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    + -> {
      3,
      x
    }
  }
};
\path (3cm,-2cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    hole
    },
  + -> {
    2,
    x
  }
};
\path (0cm,-4cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    hole
    },
  + -> {
    3,
    x
  }
};
\end{tikzpicture}
\caption{Scenario D: Deletion of Code Being Edited}
\label{fig:Scenario-D}
\end{figure}

TODO: move into Scenario B

DIAGRAM: lam x. 2 + x

DIAGRAM: (lam x. HOLE) (2 + x)

DIAGRAM: lam x. 3 + x

DIAGRAM: Merge: (lam x. HOLE) (3 + x)

A deletes code that B is editing (introduce notion of identity)

(orphans)

Figure shared between D and E that shows the deleted nodes and restore button

\subsection{Scenario E: Restoring Deleted Code}

Combine with Scenario B (moving code):
  delete
  restore
  cut/paste (i.e., move) = delete then restore
  copy structure vs copy id
  paste structure vs paste id
    (show two or three options at paste time)

DIAGRAM: 1 * 2 + 3 + HOLE

DIAGRAM: HOLE + 3 + HOLE (1 * 2)

DIAGRAM: HOLE + 3 + 1 * 2

 -- restore isn't a copy, it is a move

Multi-parent

Cycles

\subsection{Scenario F: Cycles}

@@

DIAGRAM:

@@

@@

Graph CRDT description

Mapping of syntax into the graph representation

Graph actions

holes are not explicit in the graph

\begin{theorem}[Commutativity (convergence?)]

@@

@@

\subsection{Semantic Actions}
Mapping to graph actions

Sensibility

@@

>>>>

%%%%%%%%
% Lang components

% sets
\def\E{\mathcal{E}}
\def\G{\mathcal{G}}
\def\I{\mathcal{I}}
\def\K{\mathcal{K}}
\def\U{\mathcal{U}}
\def\V{\mathcal{V}}

% constructors
\def\ExpVar{\mathsf{Exp\_var}}
\def\ExpLam{\mathsf{Exp\_lam}}
\def\ExpApp{\mathsf{Exp\_app}}
\def\ExpNum{\mathsf{Exp\_num}}
\def\ExpPlus{\mathsf{Exp\_plus}}
\def\PatVar{\mathsf{Pat\_var}}
\def\TypNum{\mathsf{Typ\_num}}
\def\TypArrow{\mathsf{Typ\_arrow}}

% indexes
\def\Root{\mathsf{Root}}
\def\LamParam{\mathsf{Lam\_param}}
\def\LamType{\mathsf{Lam\_type}}
\def\LamBody{\mathsf{Lam\_body}}
\def\AppFun{\mathsf{App\_fun}}
\def\AppArg{\mathsf{App\_arg}}
\def\PlusLeft{\mathsf{Plus\_left}}
\def\PlusRight{\mathsf{Plus\_right}}
\def\ArrowArg{\mathsf{Arrow\_arg}}
\def\ArrowResult{\mathsf{Arrow\_result}}

% relations
\DeclareMathOperator{\arity}{\text{arity}}

\newcommand{\hole}{\ensuremath{\square}} %\textcolor{violet}{\llparenthesis}}\textcolor{violet}{\rrparenthesis}}

@@

For presentational simplicity we omit param\_type in all lambdas

\subsection{Structure Editing}

We present this at two levels.
One is the underlying graph model, which
involves explicit identifiers, graph edges, indexes, and add and delete
operations.

The other is the interface a user might see,
which omits these details unless needed.

Throughout these examples the graph model is presented
as a graph with.

\subsubsection{Trees and Holes}
\label{sub:Trees and Holes}

TODO: explain holes.

\begin{figure}
\centering 
\includegraphics[scale=0.75]{graphviz/simple-tree-with-hole/1.dot}
\caption{A simple tree with a hole}
\label{fig:A simple tree with a hole}
\end{figure}
NEEDS UI EXAMPLE

Consider the tree in Figure~\ref{fig:A simple tree with a hole},
which represents the code \texttt{2 + \hole}.

Note that this tree contains three vertexes.
The first is depicted with a simple dot and is the root vertex.
The other two are labeled \texttt{plus} and \texttt{num(2)}
and represent the plus operator and the numeric literal 2, respectively.

Each vertex has a unique identifier and a list of positions
for child vertexes.
In this figure, the \texttt{plus} and \texttt{num(2)} vertexes have
identifiers 3 and 6 respectively,
the \texttt{plus} vertex has positions for its left and right children
and the \texttt{num(2)} vertex has no positions for any children.
The root vertex has the distinguished identifier 0
and has one position for its children.

Also each edge has a unique identifier. In Figure~\ref{fig:A simple tree with a hole},
these edges have idenifiers 4 and 7.  These are used
when adding and deleting edges as discussed in the
next subsubsection.

In this paper we represent variables and numeric constants
as leaves and represent their names or values as
arguments to those constructors (e.g., \texttt{2} in \texttt{num(2)}).
(In TODO:REF, we discuss how to allow edits within these.)

TODO: web page Svg ?

Declarative sentences

Outline refinement

The edge coming out of it to the TODO:word of the expression.
We call this the \emph{rooted} vertex.

\subsubsection{Delete and Add}

\begin{figure}
\centering
\begin{tikzpicture}
\path (0,0)    node {\includegraphics[scale=0.75]{graphviz/delete-and-add/1.dot}};
\path (0,-4cm) node {\includegraphics[scale=0.75]{graphviz/delete-and-add/2.dot}};
\path (0,-8cm) node {\includegraphics[scale=0.75]{graphviz/delete-and-add/3.dot}};
\end{tikzpicture}
\caption{Delete and Add}
\label{fig:Delete and Add}
\end{figure}
NEEDS UI EXAMPLE

We model edits to the code as sequences of deletions and additions of edges to this graph.
For example, consider the first graph in Figure~\ref{fig:Delete and Add}, which
represents the expression \texttt{1 + 2}.
Suppose we want to change the \texttt{2} into a \texttt{3} so the expression
becomes \textt{1 + 3}.
We model this by first deleting edge~7.
This results in the middle graph in Figure~\ref{fig:Delete and Add}, which
represents the expression \texttt{1 + \hole}.
Note that vertex~9 still exists and can be manipulated (which we will exploit in the next subsection).
Vertex~9 is just not reachable from the root.

After that deletion, we then add a new vertex to represent~\texttt{3} and an
edge to it.  This results in the bottom graph in Figure~\ref{fig:Delete and Add}.

Note that this edge has a new identifier distinct from the one before.
An edge once deleted cannot be restored.  Instead a new edge (with a new id) is created.

Explain orphans

sNote that we will omit orphans from diagrams if they are not relevant
(e.g., they are deleted and are not added back later)

\subsubsection{Move}
\begin{figure}
\centering
\begin{tikzpicture}
\path (0,0)    node {\includegraphics[scale=0.75]{graphviz/move/1.dot}};
\path (0,-5cm) node {\includegraphics[scale=0.75]{graphviz/move/2.dot}};
\path (0,-9cm) node {\includegraphics[scale=0.75]{graphviz/move/3.dot}};
\end{tikzpicture}
\caption{Move}
\label{fig:Move}
\end{figure}

\subsubsection{Cut and Paste}

cut/paste (i.e., move) = delete then add
copy structure vs copy id
paste structure vs paste id
  (show two or three options at paste time)

Restore isn't a copy, it is a move

\subsection{Collaborative Structure Editing}

\subsubsection{Editing Different Parts of the Code}

\begin{figure}
\centering
TODO: add screenshot of code text to each node
TODO: remove orphans from diagram
\begin{tikzpicture}
\path (0,0)       node {\includegraphics[scale=0.5]{graphviz/editing-different-parts-of-the-code/1.dot}};
\path (-3cm,-3cm) node {\includegraphics[scale=0.5]{graphviz/editing-different-parts-of-the-code/2A.dot}};
\path (3cm,-3cm)  node {\includegraphics[scale=0.5]{graphviz/editing-different-parts-of-the-code/2B.dot}};
\path (0cm,-6cm)  node {\includegraphics[scale=0.5]{graphviz/editing-different-parts-of-the-code/3.dot}};
\end{tikzpicture}
\caption{Editing Different Parts of the Code}
\label{fig:Editing Different Parts of the Code}
\end{figure}

commutativity

@@

\subsubsection{Editing Nested Parts of the Code}

\begin{figure}
\centering
TODO: add screenshot of code text to each node
TODO: remove orphans from diagram
TODO: annotate edges with delete and restore
\begin{tikzpicture}
\path (0,0)       node {\includegraphics[scale=0.5]{graphviz/editing-nested-parts-of-the-code/1.dot}};
\path (-3cm,-3cm) node {\includegraphics[scale=0.5]{graphviz/editing-nested-parts-of-the-code/2A.dot}};
\path (3cm,-3cm)  node {\includegraphics[scale=0.5]{graphviz/editing-nested-parts-of-the-code/2B.dot}};
\path (0cm,-6cm)  node {\includegraphics[scale=0.5]{graphviz/editing-nested-parts-of-the-code/3.dot}};
\end{tikzpicture}
\caption{Editing Nested Parts of the Code}
\label{fig:Editing Nested Parts of the Code}
\end{figure}

A deletes code that B is editing
Then A pastes it somewhere else

Explain that can continue editing orphans

@@

\subsubsection{Direct conflicts}

\begin{figure}
\centering
\begin{tikzpicture}
\path (0,0)       node {\includegraphics[scale=0.5]{graphviz/direct-conflicts/1.dot}};
\path (-3cm,-3cm) node {\includegraphics[scale=0.5]{graphviz/direct-conflicts/2A.dot}};
\path (3cm,-3cm)  node {\includegraphics[scale=0.5]{graphviz/direct-conflicts/2B.dot}};
\path (0cm,-6cm)  node {\includegraphics[scale=0.5]{graphviz/direct-conflicts/3.dot}};
\end{tikzpicture}
\caption{Direct Conflicts}
\label{fig:Direct Conflicts}
\end{figure}

Formal conflict doesn't nessisarily mean a UI conflict.
Mention resolution algorithms and that we discuss idioms
as automatic quick fixes

Multiple adds cause conflicts
(Not usually presented in the UI)

Resolve by deleting one or the other or deleting both and adding replacement

\subsubsection{Wrapping Code}

\begin{figure}
\centering
\begin{tikzpicture}
\path (0,0)       node {\includegraphics[scale=0.5]{graphviz/wrapping/1.dot}};
\path (0cm,-4cm)  node {\includegraphics[scale=0.5]{graphviz/wrapping/2.dot}};
\path (0cm,-8cm)  node {\includegraphics[scale=0.5]{graphviz/wrapping/3.dot}};
\path (0cm,-12cm) node {\includegraphics[scale=0.5]{graphviz/wrapping/4.dot}};
\path (0cm,-16cm) node {\includegraphics[scale=0.5]{graphviz/wrapping/5.dot}};
\end{tikzpicture}
\caption{Wrapping}
\label{fig:Wrapping}
\end{figure}

\subsubsection{Conflicting wrap operations}
Multi-parent

\begin{figure}
\centering
\begin{tikzpicture}
% TODO: texttt
% TODO: position the different graphs
\path (0,0)       node {\includegraphics[scale=0.5]{graphviz/conflicting-wraps/1.dot}};
\path (-3cm,-3cm) node {\includegraphics[scale=0.5]{graphviz/conflicting-wraps/2A.dot}};
\path (3cm,-3cm)  node {\includegraphics[scale=0.5]{graphviz/conflicting-wraps/2B.dot}};
\path (0cm,-6cm)  node {\includegraphics[scale=0.5]{graphviz/conflicting-wraps/3.dot}};
% \path (0cm,0cm) graph[tree layout] {
%   lam[as={$\lambda_{\texttt{x}}$}] -> {
%     x
%   }
% };
% \path (-3cm,-2cm) graph[tree layout] {
%   lam[as={$\lambda_{\texttt{x}}$}] -> {
%     + -> {
%       1,
%       x
%     }
%   }
% };
% \path (3cm,-2cm) graph[tree layout] {
%   lam[as={$\lambda_{\texttt{x}}$}] -> {
%     * -> {
%       x,
%       2
%     }
%   }
% };
% \path (0cm,-4cm) graph[tree layout] {
%   lam[as={$\lambda_{\texttt{x}}$}] -> {
%     + -> {
%       1,
%       x
%     },
%     * -> {
%       x,
%       2
%     }
%   }
% };
% \path (2cm,-8cm) graph[tree layout] {
%   lam[as={$\lambda_{\texttt{x}}$}] -> {
%     + -> {
%       1,
%       * -> {
%         x,
%         2
%       }
%     }
%   }
% };
% \path (-2cm,-8cm) graph[tree layout] {
%   lam[as={$\lambda_{\texttt{x}}$}] -> {
%     * -> {
%       + -> {
%         1,
%         x
%       },
%       2
%     }
%   }
% };
\end{tikzpicture}
\caption{Conflicting wraps}
\label{fig:Conflicting wraps}
\end{figure}

@@

\begin{figure}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    + -> {
      1,
      x
    },
    * -> {
      x,
      2
    }
  }
};
\path (2cm,-4cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    + -> {
      1,
      * -> {
        x,
        2
      }
    }
  }
};
\path (-2cm,-4cm) graph[tree layout] {
  lam[as={$\lambda_{\texttt{x}}$}] -> {
    * -> {
      + -> {
        1,
        x
      },
      2
    }
  }
};
\end{tikzpicture}
\caption{Wrap Resolution}
\label{fig:Wrap Resolution}
\end{figure}

@@

\subsubsection{Cycles}

@@

DIAGRAM: if cycles between "1 +" and "2 *" in previous diagrams

@@

% $\arraycolsep=2pt\begin{array}{llcl}
% \mathsf{Typ} & \tau & ::= & t ~\vert~ \aparr{\tau}{\tau} ~\vert~ \aall{t}{\tau} ~\vert~ \arec{t}{\tau} ~\vert~ \aprod{\labelset}{\mapschema{\tau}{i}{\labelset}} ~\vert~ \asum{\labelset}{\mapschema{\tau}{i}{\labelset}}\\
% \mathsf{Exp} & e & ::= & x ~\vert~ \aelam{\tau}{x}{e} ~\vert~ \aeap{e}{e} ~\vert~ \aetlam{t}{e} ~\vert~ \aetap{e}{\tau} ~\vert~ \aefold{e} ~\vert~ \aetpl{\labelset}{\mapschema{e}{i}{\labelset}} ~\vert~  \aein{\ell}{e} \\
% & & \vert & \aematchwith{n}{e}{\seqschemaX{r}}\\
% \mathsf{Rule} & r & ::= & \aematchrule{p}{e}\\
% \mathsf{Pat} & p & ::= & x  ~\vert~ \aewildp ~\vert~ \aefoldp{p} ~\vert~ \aetplp{\labelset}{\mapschema{p}{i}{\labelset}} ~\vert~ \aeinjp{\ell}{p}
% \end{array}$

\begin{figure}
\[
\arraycolsep=0pt
\begin{array}{rlll}
 p    & {}\in Pat & {}\mathrel{::=}{} & x \mid \_ \\
 e    & {}\in Exp & {}\mathrel{::=}{} & \lambda p\mathord{:}\tau.e \mid e~e \mid e + e \mid n \mid \_ \\
 \tau & {}\in Typ & {}\mathrel{::=}{} & \tau \rightarrow \tau \mid Num \mid \_
\end{array}
\]
\caption{Formal syntax (as a grammar) for the lambda calculus that we are considering.}
\label{fig:formal-syntax}
\end{figure}

Turn that into a constructor set K + index set I + arity function (has type: K -> $\wp$(I)).

$\K = \{ \Root, \ExpVar, \ExpLam, \ExpApp, \ExpNum, \ExpPlus,$

$\qquad\ \ \PatVar, \TypNum, \TypArrow \}$

$\I = \{ \Root, \LamParam, \LamType, \LamBody, \AppFun, \AppArg,$

$\qquad\ \ \PlusLeft, \PlusRight, \ArrowArg, \ArrowResult \}$

\vspace{2ex}

$\arity : \K \rightarrow \wp(\I)$

\vspace{2ex}

\begin{tabular}{ll}
    $\K$ & $\wp(\I)$ \\
    \hline
    $\Root$ & $\{ \Root \}$ \\
    $\ExpLam$ & $\{ \LamParam, \LamType, \LamBody \}$ \\
    $\ExpApp$ & $\{ \AppFun, \AppArg \}$ \\
    $\ExpPlus$ & $\{ \PlusLeft, \PlusRight \}$ \\
\end{tabular}

@@

\subsubsection{Graphs}

Graph is a function from edges to edge states.

$\G : \E \rightarrow \{ \bot, +, - \}$

An edge is a uuid * source vertex * index * target vertex.  (We are assuming you can generate globally unique IDs.)

\def\e{\varepsilon}

An edge $\e \in \E = \U \times \V \times \I \times \V$ is ...

A vertex is a pair of a uuid and a constructor. (uuid + constructor is the identity)

$\V = \U \times \K$

An edge state is bottom + - . Define the total ordering (forms a lattice).

Define $\sqcup_{edge state}$ (i.e. join)

lattice ordering $\bot < + < -$ LUB e.g., join A B = C

You can map from syntax to graph by selecting
unique IDs (for both edges and vertexes).
    - holes are not explicit in the graph

\subsubsection{Graph Actions}
A graph action is a +E or -E.

Graph action semantics is a transition system
between graphs. (join with whatever it was before)

G --action--> G'

Discuss uniqueness of uuid

\begin{theorem}[Commutativity]

G  -- a1 a2 --> G'

------------------

G' -- a2 a1 --> G

@@

editor states:

@@

----------------

User must supply edge selection for multi-parent

\subsection{Edit Actions}
Mapping to graph actions

Sensibility -- mapping back from graphs
to well-typed states that fixes holes

\section{TODO}
TODO: leaves

TODO: our model supports treating these as a cons-list of characters

@@

\subsection{Variable names, strings, and numbers}

NOTE: we could implement each digit as a separate characters

GUI for string conflicts: use popups
